\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-22 paper.qmd
Removed duplicated figures

umap_pbmc <- umap_pbmc_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2)) +
  geom_point(alpha=0.3, color = "#000000") 
```

<!-- UMAP parameters suggested by the author (n_neighbors = 30, min_dist = 0.3)-->
```{r}
#| fig-cap: "Reproduction of plot published in Chen et al. (2023) showing a $2\\text{-}D$ layout from UMAP applied for the PBMC3k dataset. The (hyper-)parameter settings, beyond the defaults, are 30 nearest neighbors and minimum distance equal to 0.3. We use our model-in-the-data-space to assess whether this is an accurate representation of structure present in the high-dimensional data, or if it is misleading."
#| label: fig-umap-author
#| out-height: 30%
#| fig-pos: H

@@

<!--bin1 = 30, bin2 = 29, b = 870, non_empty = 135-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in $2\\text{-}D$ with UMAP, and (b) $p\\text{-}D$ model error in $2\\text{-}D$. The $2\\text{-}D$ model shows three well-separated distant clusters. The $p\\text{-}D$ model errors are distributed along clusters."
#| label: fig-model-pbmc-author
#| fig-pos: H

@@

tsne_pbmc <- tsne_pbmc_scaled |>
  ggplot(aes(x = tSNE1,
             y = tSNE2)) +
  geom_point(alpha=0.3, color = "#000000")
```

<!--tSNE with perplexity:30-->
```{r}
#| echo: false
#| fig-cap: "The $2\\text{-}D$ layout from tSNE applied for the PBMC3k dataset. The default (hyper-)parameter setting is a perplexity $30$. We use our model-in-the-data-space to assess whether this is an accurate representation of structure present in the high-dimensional data, or if it is misleading."
#| label: fig-tsne-suggest
#| fig-pos: H
#| out-width: 40%

>>>>

# umap_pbmc <- umap_pbmc_scaled |>
#   ggplot(aes(x = UMAP1,
#              y = UMAP2)) +
#   geom_point(alpha=0.3, color = "#000000") 
```

<!-- UMAP parameters suggested by the author (n_neighbors = 30, min_dist = 0.3)-->
```{r}
#| fig-cap: "Reproduction of plot published in Chen et al. (2023) showing a $2\\text{-}D$ layout from UMAP applied for the PBMC3k dataset. The (hyper-)parameter settings, beyond the defaults, are 30 nearest neighbors and minimum distance equal to 0.3. We use our model-in-the-data-space to assess whether this is an accurate representation of structure present in the high-dimensional data, or if it is misleading."
#| label: fig-umap-author
#| out-height: 30%
#| fig-pos: H
#| eval: false

@@

<!--bin1 = 30, bin2 = 29, b = 870, non_empty = 135-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in $2\\text{-}D$ with UMAP, and (b) $p\\text{-}D$ model error in $2\\text{-}D$. The $2\\text{-}D$ model shows three well-separated distant clusters. The $p\\text{-}D$ model errors are distributed along clusters."
#| label: fig-model-pbmc-author
#| fig-pos: H
#| eval: false

@@

# tsne_pbmc <- tsne_pbmc_scaled |>
#   ggplot(aes(x = tSNE1,
#              y = tSNE2)) +
#   geom_point(alpha=0.3, color = "#000000")
```

<!--tSNE with perplexity:30-->
```{r}
#| echo: false
#| fig-cap: "The $2\\text{-}D$ layout from tSNE applied for the PBMC3k dataset. The default (hyper-)parameter setting is a perplexity $30$. We use our model-in-the-data-space to assess whether this is an accurate representation of structure present in the high-dimensional data, or if it is misleading."
#| label: fig-tsne-suggest
#| fig-pos: H
#| out-width: 40%
#| eval: false

### 2024-08-22 paper.tex
Updated Figure 2
Removed duplicated figures

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
NLDR is commonly used to display clusters, and help to verify the
results. For example, \citet{chen2023} illustrates the use of UMAP to
identify clusters in Human Peripheral Blood Mononuclear Cells (PBMC3k).
Figure~\ref{fig-umap-author} is a reproduction of the published plot. It
shows three well-separated clusters with big separations.

\includegraphics[width=1\textwidth,height=0.3\textheight]{paper_files/figure-pdf/fig-umap-author-1.pdf}

\caption{\label{fig-umap-author}Reproduction of plot published in Chen
et al.~(2023) showing a \(2\text{-}D\) layout from UMAP applied for the
PBMC3k dataset. The (hyper-)parameter settings, beyond the defaults, are
30 nearest neighbors and minimum distance equal to 0.3. We use our
model-in-the-data-space to assess whether this is an accurate
representation of structure present in the high-dimensional data, or if
it is misleading.}

To determine whether the UMAP representation with the (hyper-)parameter
choice suggested by \citet{chen2023} preserves the original data
structure, we visualize the model constructed with UMAP overlaid on the
\(9\text{-}D\) data. The Figure~\ref{fig-umap-author} shows three
well-separated clusters. However, as shown in Figure~\ref{fig-pbmc1-sc},
there is no big separation between three clusters in \(9\text{-}D\).
Therefore, the suggested UMAP representation
(Figure~\ref{fig-umap-author}) does not accurately represent the
structure(s) present in PBMC3k dataset. But, when visualizing the
model-in-th-data-space some unobserved structures can be seen. Some
clusters have non-linear continuity patterns and high-density patches
(Figure~\ref{fig-pbmc1-sc}).

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-model-pbmc-author-1.pdf}

\caption{\label{fig-model-pbmc-author}(a) Model generated in
\(2\text{-}D\) with UMAP, and (b) \(p\text{-}D\) model error in
\(2\text{-}D\). The \(2\text{-}D\) model shows three well-separated
distant clusters. The \(p\text{-}D\) model errors are distributed along
clusters.}

@@

As shown in Figure~\ref{fig-tsne-suggest}, there are three
well-separated clusters, although the separation between the clusters
are small. Additionally, non-linear structures can also be observed
within the clusters (Figure~\ref{fig-tsne-suggest}). This demonstrates
that tSNE accurately captures the data structure of the PBMC3k dataset,
which UMAP did not achieve.

\includegraphics[width=0.4\textwidth,height=\textheight]{paper_files/figure-pdf/fig-tsne-suggest-1.pdf}

\caption{\label{fig-tsne-suggest}The \(2\text{-}D\) layout from tSNE
applied for the PBMC3k dataset. The default (hyper-)parameter setting is
a perplexity \(30\). We use our model-in-the-data-space to assess
whether this is an accurate representation of structure present in the
high-dimensional data, or if it is misleading.}

@@

\begin{center}
\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/unnamed-chunk-60-1.pdf}
\end{center}

>>>>

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
NLDR is commonly used to display clusters, and help to verify the
results. For example, \citet{chen2023} illustrates the use of UMAP to
identify clusters in Human Peripheral Blood Mononuclear Cells (PBMC3k).
\textbf{?@fig-umap-author} is a reproduction of the published plot. It
shows three well-separated clusters with big separations.

To determine whether the UMAP representation with the (hyper-)parameter
choice suggested by \citet{chen2023} preserves the original data
structure, we visualize the model constructed with UMAP overlaid on the
\(9\text{-}D\) data. The \textbf{?@fig-umap-author} shows three
well-separated clusters. However, as shown in Figure~\ref{fig-pbmc1-sc},
there is no big separation between three clusters in \(9\text{-}D\).
Therefore, the suggested UMAP representation
(\textbf{?@fig-umap-author}) does not accurately represent the
structure(s) present in PBMC3k dataset. But, when visualizing the
model-in-th-data-space some unobserved structures can be seen. Some
clusters have non-linear continuity patterns and high-density patches
(Figure~\ref{fig-pbmc1-sc}).

@@

As shown in \textbf{?@fig-tsne-suggest}, there are three well-separated
clusters, although the separation between the clusters are small.
Additionally, non-linear structures can also be observed within the
clusters (\textbf{?@fig-tsne-suggest}). This demonstrates that tSNE
accurately captures the data structure of the PBMC3k dataset, which UMAP
did not achieve.

@@

\begin{center}
\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/unnamed-chunk-61-1.pdf}
\end{center}
